---
title: "team"
author: "Armando Castillo"
date: "11/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate) 
library(ggplot2)
library(scales)
library(tidyverse)
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
GLC <- readxl::read_xlsx("FRPP_GLC_-_United_StatesSep292020.xlsx")
```

1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?

```{r}
acc$DAY_WEEK = as.factor(acc$DAY_WEEK)
levels(acc$DAY_WEEK) <-  c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
acc%>%
  group_by(DAY_WEEK)%>%
  summarise(total = n(), .groups = 'drop')
acc%>%
  group_by(DAY_WEEK)%>%
  summarise(total = n(), .groups = 'drop')%>%
  ggplot(aes(x = DAY_WEEK, y = total))+
  geom_bar(stat = 'identity') +
  ggtitle('Accidents per Day of Week')+
  xlab('Day of Week')
```

Answer:  Friday through sundays has the most accidents. With Saturday at number 1, then Friday and then Sunday at third. This is most likley there are more drivers and more night driving.


2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)

```{r}
fatal <- person%>%
  filter(INJ_SEV == 4)
head(fatal)
```

3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)
```{r}
make <- left_join(acc, fatal, by="ST_CASE")
make <- make%>%
  rename(STATE = STATE.x)%>%
  select(STATE, MAKE)
make <- na.omit(make)
make <- make%>%
  group_by(STATE, MAKE)%>%
  summarise(total = n(), .groups = 'drop')%>%
  group_by(STATE)%>%
  top_n(1, total)
    
head(make)

```

4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)
```{r}
USA <- readxl::read_xlsx("FRPP_GLC_-_United_StatesSep292020.xlsx")
View(USA)
states <- map_data("state")
states$region = toupper(states$region)
makeModel <- left_join(acc, fatal, by="ST_CASE")
makeModel <- makeModel%>%
  rename(STATE = STATE.x)%>%
  select(STATE, MAK_MOD)
makeModel <- na.omit(makeModel)
makeModel <- makeModel%>%
  group_by(STATE, MAK_MOD)%>%
  summarise(total = n(), .groups = 'drop')%>%
  group_by(STATE)%>%
  top_n(1, total)

```

5. Join the accident and person table (work out which variable(s) to use)
```{r message=FALSE, warning=FALSE}
ACC_PER<-acc%>%
  left_join(person,by="ST_CASE")%>%
  distinct(ST_CASE, .keep_all = TRUE)
```

6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.


7. Now plot a choropleth map of the number of deaths on a county level.
```{r}
death<-acc %>% 
  group_by(STATE, COUNTY) %>% 
  summarise(deaths=sum(FATALS))
county <- map_data("county")
GLC <- GLC %>% 
  rename(STATE=`State Code`, COUNTY=`County Code`) %>% 
  mutate(STATE=as.numeric(STATE), COUNTY=as.numeric(COUNTY))
Map2 <- death %>% 
  inner_join(GLC, by=c("STATE", "COUNTY")) %>% 
  select(deaths, `State Name`, `County Name`) %>% 
  distinct() %>%
  rename(region=`State Name`, subregion=`County Name`) %>% 
  mutate(region=tolower(region), subregion=tolower(subregion))%>% 
  right_join(county, by=c("region", "subregion"))
Map2 %>% 
  ggplot(aes(x = long, y = lat, fill=deaths),color = 'white') +
  geom_polygon(aes(group=group))

```

8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.

```{r}
season <- fatal %>%
  filter(MONTH %in% c(6,7,8,9,12,1,2,3)) %>%
  mutate(season=ifelse(MONTH %in% c(6,7,8,9), 'summer', 'winter')) %>%
  mutate(dead = ifelse(DOA %in% c(7,8), 1 ,0)) %>%
  group_by(STATE, season) %>%
  summarize(deaths=sum(dead)) %>% 
  left_join(GLC, by='STATE')
View(season)
ggplot(season, aes(x=' STATE NAME', weight=deaths, fill=season)) + 
  geom_bar(position='fill') + 
  coord_flip() +
  ylab('Proportion of Deaths') +
  xlab('State') + 
  ggtitle('Deaths in Summer and Winter by State')
```

Answer: There are more accidents in the summer time than the winter. But looks like areas that already had accidents during summer also have them during winter. So there are clear areas that are prone to accidents.




